Artificial ecosystem selection reveals relationships between microbiome composition and ecosystem function

### 2022-04-07 Manuscript/manuscript.Rmd
Update manuscript

---
title: Artificial ecosystem selection to deduce the mapping between microbial community structure and ecosystem function
author: "Andrew H. Morris and Brendan J. M. Bohannan"
output:
  word_document:
    reference_docx: reference.docx
editor_options:
  chunk_output_type: inline
---

```{r options, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r initialize_packages}
library(tidyverse)
library(captioner)
library(broom)
library(broman)
library(morris)
#library(metafor)
```
```{r load_data}
heritability <- read_tsv('../Output/heritability.tsv')
fluxes <- read_tsv('../Output/fluxes.tsv')

@@

@@

```{r figure_captions, include=FALSE}

figures <- captioner()
figures("response", "Deviance of CH~4~ oxidation rate for the Positive selection treatment relative to the Neutral selection treatment. Deviance is calculated as the mean of P within each passage minus the mean of N within each passage. Errorbars are the standard error of the difference between P and N for each passage. The best fit line is an ordinary least squares fit taking into account the standard errors of the deviations. The gray ribbon is the 95% confidence interval for the regression.")
figures("herit", "Regression of offspring CH~4~ flux on mid-parent CH~4~ flux. Mid-parent is the mean of jars selected to inoculate the next passage. The mid-offsprings are the mean of all twelve jars produced in one passage. The best fit line is an ordinary least squares regression line and the gray ribbon is the 95% confidence interval for the regression coefficient.")

@@

One of the fundamental questions in Ecology is To what extent does biodiversity
drive variation in ecosystem function? Variation in abiotic conditions
regulates the rate of ecosystem processes, but it remains unclear to what
extent variation in community structure regulates the rate of ecosystem
processes. Many ecosystem functions are mediated by organisms. In particular,
microorganisms such as bacteria mediate the major biogeochemical cycles such
as the carbon and nitrogen cycles. Despite the central role of bacteria in
performing these processes, we have not found a consistent relationship between
microbial community structure and the rate of an ecosystem process. In order
to understand the importance of biodiversity to ecosystem functioning and to
better predict and manage specific ecosystem processes, we need to determine
the relationship between microbial community structure and ecosystem function.

In macroorganismal communities, there appears to be a relationship between the
diversity of a community and the resilience of ecosystem functions to
environmental perturbations (Tilman, ????... Hooper). Typically, this is the
diversity (or more specifically, the richness) of a particular functional
group, such as photosynthetic plants. Similar studies in microbial communities
rarely find a relationship between diversity and function (Graham, ). However,
these studies typically define diversity much more broadly than plant diversity
studies by looking at the phylogenetic diversity of the entire bacterial and
archaeal community. Studies that focus more narrowly on a particular ecosystem
process and a functional group involved in that process still rarely find a
significant correlation between the abundance of a functional group and the
rate of the corresponding process (Rocca, ????). For certain ecosystem
functions in certain ecosystems, these relationships can be informative. For
example, the abundance of a single methanogenic species (*G. species*) is
correlated with methane emissions in a permafrost ecosystem (cite). In
addition, bacterial and archaeal community composition is correlated with
methane emissions in forests and pastures of the Brazilian Amazon (Meyer et al.
2020).  Therefore, there appears to be some indication that microbial community
structure drives variation in methane cycling in soils.

To further investigate the relationship between microbial community structure
and the rate of methane cycling in soils, we want to determine which organisms
are important for regulating methane cycling. This ecological question is
analogous to the problem of genotype-phenotype mapping in quantitative genetics
(Morris 2020). Variation in an organismal trait is partially determined by
environmental conditions and partially determined by genetic variation. One goal
of quantitative genetics is to determine the proportion of the variation
attributable to genetics and to identify which alleles regulate the phenotypic
trait. These goals are commonly achieved through artificial selection
experiments and association mapping studies, such as genome-wide association
studies. To determine the mapping between microbial community structure and
ecosystem function, we applied these approaches of genotype-phenotype mapping
to whole microbial communities. In this case, the "alleles" are different
microbial taxa and the "phenotype" is the rate of ecosystem function at the
whole-ecosystem level.

To identify communities of bacteria and archaea that collectively perform a
high rate of methane oxidation, we performed an artificial ecosystem selection
experiment  (Swenson et al. 2000).  We generated twenty-four soil ecosystems in the lab using a sterilized
potting mix as the substrate. We then inoculated these ecosystems with a small
amount of living soil to generate variation in community structure across the
twenty-four ecosystems. These ecosystems were then assigned to one of two
selection treatments: neutral selection or positive selection. Each ecosystem
was maintained at 1000 ppm CH~4~ over several weeks to allow for colonization and
growth of the microbial community and to enrich for methane oxidizers. We then
determined the methane oxidation rate for each ecosystem. To impose selection,
we selected three jars from the positive treatment with the highest rate of
methane oxidation, homogenized them, and used that soil to inoculate the next
set of jars. For the neutral treatment, we selected three jars at random to
inoculate the next set of jars. This process continued for five passages. At
the end of the experiment, we extracted DNA from soils in the second and fifth
passage for both the neutral and positive treatments. By comparing the response
to selection within the community between the two treatments, we can identify
microbial species that contribute to variation in ecosystem methane emissions.

This approach is powerful because by using a common soil substrate and
maintaining a constant headspace concentration of methane, we can eliminate
much of the environmental variation that would be present in an observational
study. In addition, by applying selection on methane oxidation rate
in the positive treatment, we can enrich for taxa involved in methane cycling
making it easier to perform association mapping between microbial taxa and
ecosystem methane emissions. In this paper, we address whether there is a
response to selection on methane oxidation rate at the whole ecosystem level. We
then quantify the amount of variation in methane oxidation attributable to
variation in community composition within our laboratory environment. Finally,
we identify microbial markers of soil methane cycling using association mapping
in order to deduce the mapping between microbial community structure and
ecosystem function.

(Goodnight et al. 1997, Williams and Lenton
2007)

# Materials & Methods

## Selection experiment
We performed an artificial ecosystem selection experiment similar to Swenson et
al. (2000) by passaging replicate soil microbial communities. The trait we
selected on was CH~4~ oxidation rate. Our experiment had two selection lines
with twelve jars each for a total of twenty-four jars per passage. One line was
a positive selection line where the two or three jars with the highest CH~4~
oxidation rate were chosen to inoculate the next set of positive jars. The
other line was a neutral selection line where an equal number of jars as the
positive line were chosen at random to inoculate the next set of neutral jars.
The number of jars chosen was based on the distribution of fluxes among the
positive jars, i.e., we chose the top three jars unless only two jars had
considerably greater oxidation rates based on visual inspection of histograms.
The experiment was carried out over five passages.

The initial soil microbial community was sampled from the top 10 cm of an
upland mineral soil under a deciduous forest ecosystem near the University of
Oregon campus in Eugene, OR, USA. Incubations were performed in 500 mL mason
jars (Ball Corporation, Broomfield, Colorado, USA) with rubber septa installed
in the lids. Each jar was sterilized with 70% ethanol to which was added 45 g
of autoclaved potting mix, 5 g of living soil, and 3.5 mL of sterile deionized
water to bring them to <mark> 60%??? of field capacity</mark>. The jars were
then homogenized, capped, and injected with 4.3 mL of 99% CH~4~ to bring the
headspace concentration to approximately 1000 ppm CH~4~. To create the two
treatment groups, twenty-four jars were created in an identical manner and then
randomly assigned to either the positive or neutral selection line. Jars were
flushed and respiked twice per week to maintain elevated CH~4~ concentrations
and were incubated at ambient temperature for approximately three weeks.
Methane oxidation rates were determined at the end of the incubation period.
These values were then used to determine the jars selected to inoculate the
next set of jars.

## Methane oxidation rate measurements

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from
each jar immediately after spiking and then at time points 3, 6, 24, and 48
hours for a 5-point curve. Samples were immediately injected into a
<mark>???</mark> gas chromatograph fitted with an electron capture detector
(company, city, state, USA) to determine the headspace CH~4~ concentration.
Fluxes were calculated from a first-order exponential decay function as decay
constant `k` with units day^-1^. Oxidation rates are presented as the additive
inverse of `k` (i.e., `-k`) so that a more positive value represents a faster
oxidation rate. The jars selected for the positive treatment in passage 2 had
the lowest methane oxidation rate of the twelve jars due to a calculation error
in the methane oxidation rate. All other passages correctly used the top three jars.

A subsample of soil from the starting inoculum and from every jar in passages 2
and 5 was collected and stored at -80$^\circ$C for later DNA extraction.
Soil DNA was extracted from a 0.25 g subsample of soil. Negative controls were extracted
from autoclaved potting mix and DNase-free water. Extractions were performed
using the DNeasy PowerSoil kit (QIAGEN, Hilden, Germany) and quantified using
Qubit (company, city, state). In order to estimate the diversity and relative
abundance of the bacterial and archaeal taxa in our soil ecosystems, we
sequenced the V3-V4 region of the 16S rRNA gene using the Earth Microbiome
Project protocol and primers<mark>???</mark>. Sequencing was performed on the
Illumina NovaSeq 6000 with paired-end 150 bp reads.

Bioinformatic analyses were conducted using the `QIIME2`
bioinformatics platform (Bolyen et al. 2019). Raw sequencing reads were
demultiplexed using ??? (citation), denoised using `DADA2` (citation).
Taxonomic assignment for the 16S reads was performed using ??? based on the
Silva database. All subsequent tatistical analyses were performed in the `R`
statistical programming environment (R Core Team 2020). Richness estimates were
made using the `breakaway` package and diversity indices using the `DivNet`
package (Willis et al. ???).

### Response to selection

To test whether there was a significant change in CH~4~ oxidation rate (k) as a
response to selection, we tested a difference in slopes between the positive
and neutral selection lines. The CH~4~ oxidation rates were strongly right
skewed with most values close to zero and few large, positive values. This
resulted in residuals that did not meet the assumptions of constant variance
and normal distribution. Therefore, methane oxidation rates were log~10~
transformed to better meet the assumptions of a linear model and to make
figures easier to interpret. First, we tested the effect of treatment by
fitting two nested models with and without treatment using `lm` in `R`. The
model formulation for the full model was `flux ~ passage * treatment` and the
model formulation for the reduced model was `flux ~ passage`. We compared these
models using the likelihood ratio test with the `anova` function in `R`. We
then fit the full model to determine the slope of the positive line, which
represents the change in methane oxidation rate per passage as a response to
selection.

### Ecosystem heritability

To estimate the proportion of variance in ecosystem function due to additive
genetic variance of the microbial communtiy, we applied the Breeder's equation
to the results of our selection experiment. the heritability of
CH~4~ oxidation rate as an ecosystem property, we calculated narrow-sense
heritability ($h^2$) as the regression of offspring on mid-parent (Goodnight
2000, Falconer 1997). The mid-parent was the mean for all three selected jars
and the offspring was the mean for all jars produced by those parents.

In addition, we calculated the inbreeding coefficient, $F$, as:

$$
F = 1 - \left(1 - \frac{1}{2N} \right) ^{T}
$$

where $N$ is the population size and $T$ is the number of generations (Goodnight 2000 or Falconer 1997).

### Taxa differentially abundant in the positive selection treatment

To identify taxa that responded to selection on CH~4~ oxidation rate, we tested
differential abundance using a beta-binomial model (Martin et al. 2020). We
compared taxa in passage 5 between the positive and neutral treatment to
identify taxa that

Possible tests:

- Random forest or gradient boosting to predict Positive vs. Neutral.

## Response to selection

We observed a response to selection on soil methane oxidation rate at the whole
ecosystem level (`r figures("response", display="cite")`).
At the start of the experiment, the Positive
treatment had a mean methane oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`%
lower than the
Neutral treatment (SE = 
`r myround(tidy(response_model)[[3, 3]], 2)`, t = 
`r myround(tidy(response_model)[[3, 4]], 2)`, p = 
`r myround(tidy(response_model)[[3, 5]], 2)`). There was no change in methane oxidation rate of the Neutral
treatment over the five passages (slope = 
`r myround(tidy(response_model)[[2, 2]], 2)`, SE = 
`r myround(tidy(response_model)[[2, 3]], 2)`, t = 
`r myround(tidy(response_model)[[2, 4]], 2)`, p = 
`r myround(tidy(response_model)[[2, 5]], 2)`). By contrast, the Positive treatment had a
`r round(per_slope)`% increase in methane
oxidation rate per passage (slope = 
`r myround(tidy(response_model)[[4, 2]], 2)`, SE = 
`r myround(tidy(response_model)[[4, 3]], 2)`, t =
`r myround(tidy(response_model)[[4, 4]], 2)`, p = 
`r myround(tidy(response_model)[[4, 5]], 2)`).

```{r response_figure, fig.width=4, fig.height=3, fig.cap=figures("response")}
ratio_log10 <- calc_plot_ratios(fluxes$passage, fluxes$log_ch4)

plot_flux(fluxes, passage, log_ch4, ratio_log10, log10 = TRUE)

@@

## Ecosystem heritability

Here, I fit two models. The first is simply offspring flux on selected parent
flux, ignoring treatment. There is no correlation. Next, I added treatment and the interaction between treatment and parental
flux. Here, there is a significant correlation between parent and offspring for
both the Neutral treatment `r print_model(herit_model, 2)` and the Positive
treatment `r print_model(herit_model, 4)`. Notably, the sign of the effect is
reversed: negative for the Neutral treatment and positive for the Positive
treatment.

The negative heritability of methane oxidation rate in the Neutral selection
treatment may result from selection on alternative characters in those
populations. The strong, positive heritability in the Positive treatment ... .
It seems strange that the Methane oxidation rate was heritable between
consecutive passages (`r figures("herit", display="cite")`). Specifically, the
narrow-sense heritability, calculated as the slope of the mid-parent and
mid-offspring, was $h^2$ =  $\pm$  (t = , p = ). With an inbreeding coefficient of ,

```{r herit_fig, fig.width=5, fig.height=3, fig.cap=figures("herit")}
plot_herit(heritability)
```

>>>>

---
title: Artificial ecosystem selection reveals relationships between microbiome composition and ecosystem function
author: "Andrew H. Morris^1^* and Brendan J. M. Bohannan^1^"
output:
  word_document:
    reference_docx: reference.docx
bibliography: bibliography.bib
csl: vancouver.csl
---

^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA

\* Correspondence: Andrew H. Morris, amorris3@uoregon.edu

Competing Interests: We declare we have no competing interests.

```{r options, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

```{r initialize-packages}
library(tidyverse)
library(broom)
library(broman)
library(morris)
library(phyloseq)
library(ggpubr)
library(knitr)
library(corncob)
library(ggthemes)
library(microViz)
library(captioner)
figures <- captioner()
tables <- captioner(prefix = "Table")
```

```{r load-data}
heritability <- read_tsv('../Output/heritability.tsv')
fluxes <- read_tsv('../Output/fluxes.tsv')

@@

load(file = '../Output/richness_model_passage.Rdata')
load(file = '../Output/richness_model_treat5.Rdata')
load(file = '../Output/richness_model_ch4.Rdata')
load(file = '../Output/phyloseq.RData')
load(file = '../Output/beta_model.Rdata')
load(file = '../Output/da_corncob.Rdata')
load(file = '../Output/dt_data.Rdata')

@@

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
any ecosystem function. This challenge remains because microbiomes are often
complex, it is difficult to know a priori which microbiome members limit the
rate of an ecosystem function, and it is usually challenging to manipulate
microbiome composition directly. We propose that artificial ecosystem selection
can overcome these challenges. Artificial ecosystem selection involves creating
replicate ecosystems, selecting replicates with a desired ecosystem-level
trait, and using the microbiomes from the selected ecosystems to inoculate a
new set of ecosystems. We performed artificial ecosystem selection on replicate
soil microcosms to select for increased soil methane oxidation rate. We
observed a strong response of methane oxidation rate to artificial ecosystem
selection with a 51% increase in methane oxidation rate per round over five
rounds of selection. This suggests that a large portion of the variation in
methane oxidation rate in these soils can be attributed to microbiome
variation. We also found that selection did not enrich for known methane
oxidizers; rather microbes that were enriched after selection included 24
families such as Puniceicoccaceae, Elsteraceae, and Cytophagaceae. This result
is in contrast to the typical assumption that the rate of an ecosystem function
is limited by the final step in the associated microbial pathway. Our study
demonstrates that artificial ecosystem selection is a useful strategy for
revealing relationships between microbiome composition and ecosystem function.

Microbiomes mediate a variety of important functions in ecosystems, and there
is great interest in understanding how attributes of microbiomes may influence
variation in ecosystem functions [@crowther2019]. Biodiversity-ecosystem
function relationships have been described for a variety of macroorganismal
communities. For example, plant species richness correlates with productivity
and marine community diversity correlates with functions such as production,
consumption, and nutrient release
[@cardinale2006; @hooper2012; @gamfeldt2015]. While there is evidence that
variation in the microbiome contributes to variation in ecosystem functions,
describing this relationship has been elusive [@rocca2015; @graham2016].

One reason for this difficulty is that it is challenging to directly manipulate
microbiome composition and diversity the way we can for macroorganismal
communities. Instead, we must use genetics through marker gene or metagenomic
sequencing to characterize the microbiome.  However, studies that measure the
taxonomic or functional diversity of a microbiome through these approaches
rarely find a correlation between diversity and function [@graham2016]. Another
approach is to quantify cells or genes using  quantitative
PCR. Studies such as this still rarely find a correlation between the abundance
of a functional group and the rate of the corresponding process [@rocca2015].
As a result, it remains unclear how and when microbiome variation contributes
to variation in the rate of ecosystem functions.

Another likely reason that microbiome-ecosystem function relationships have been
difficult to document is that the most common approaches for looking for a
relationship requires some prior knowledge of the likely causal links between
microbiomes and ecosystem functions. For example, many studies have looked for
relationships with the abundance or diversity of a "marker gene" for a
particular function. Marker genes are microbial genes that code for an enzyme or
enzyme-subunit known to be involved in a particular function. But this approach
assumes that one can identify the "right" marker gene a priori which, given
how little is understood about microbial diversity, seems highly unlikely. What
is needed are approaches to microbiome-ecosystem
function mapping that do not require this degree of prior knowledge.

Yet another likely reason that microbiome-ecosystem function relationships have
been difficult to document is that microbiomes can be related to ecosystem
functions in two related but distinct ways that historically have been difficult
to separate. One way is that microbiomes may simply be conduits through which
the environment drives function. That is to say that the environmental
conditions completely regulate the attributes of the microbiome and a shift in
the environment causes a shift in microbiome attributes, such as the abundance
of a microbial functional group. This shift results in a change in the rate of
an ecosystem function. In this case, microbiome attributes would be
statistically associated with variation in ecosystem function, but also strongly
covary with environmental conditions. Therefore, there would be no association
between microbiomes and ecosystem functions independent of the environmental
conditions. Under this scenario, knowing anything about the microbiome would not
necessarily provide you with information regarding the causes of the change in
ecosystem function or improve one's ability to predict changes in ecosystem
function. Furthermore, attempting to alter ecosystem function by altering the
microbiome without changing the underlying environmental conditions would likely
be fruitless.

Another possibility is that the microbiome itself alters the rate of ecosystem
function independent of the environment. In this case, one cannot predict the
rate of ecosystem function without understanding changes in the microbiome. In
addition, managing the microbiome through inoculation or selective antibiotics
would potentially be an effective approach for enhancing or mitigating certain
changes to ecosystem functions. These two scenarios are not mutually
exclusive, nor are they likely unique to microbes, but methods to determine the
relative important of these two scenarios have not until recently been employed.

The problem of connecting microbiome attributes to the rate of an ecosystem
function is analogous to the problem of connecting genomic variation to
phenotypic traits in organisms, and solutions to this problem may be found by
exploring this analogy [@morris2020]. For example, one approach to the problem of
genotype-phenotype mapping is to perform artificial selection on a trait of
interest and identify genes that respond to selection. Such genes are likely to
be causally linked to the trait under selection. An analogous approach for
exploring microbiome-ecosystem function relationships is gaining acceptance
among microbiologists [@swenson2000; @panke-buisse2015]. Microbiomes have been
shown to respond to selection at the ecosystem-trait level [@swenson2000],
and some ecosystem-scale microbiome traits have been shown to be transferable
through microbiome inoculations [@panke-buisse2015]. Applying artificial
selection to whole microbiomes could be a novel approach for quantifying how
much of the variation in the rate of an ecosystem function could be attributed
independently to microbiome variation [@goodnight2000]. In addition, this
approach could be used to identify microbiome attributes, such as genes or taxa,
associated with the rate of an ecosystem function, although, to my knowledge
this has not previously been accomplished [@arias-sanchez2019; @morris2020]. 
Such an approach has the potential to expand our ability to model
microbial ecosystem-function relationships more accurately and allow us to
manage ecosystems for particular outcomes [@fuhrman2009; @schimel1998]. Here I
use this approach to estimate how much of the variation in ecosystem function
(the flux of CH~4~ gas between the soil and atmosphere) can be attributed to
variation in the microbiome and to identify microbial markers associated with
this ecosystem function.

Soil CH~4~ oxidation is a suitable function for this study because there is
evidence that it may vary with microbiome attributes. For example, variation in
CH~4~ emissions in arctic permafrost is correlated with the transcriptional
activity of certain methanogens [@freitag2009; @freitag2010]. In addition,
soil microbiome phylogenetic variation is a strong predictor of CH~4~ oxidation in
forests and pastures of the Brazilian Amazon [@meyer2020]. Finally,
methanogenesis and methanotrophy are two of the most deeply conserved microbial
physiologies and are represented in a narrow range of taxa and so the taxonomic
composition of the microbiome is more likely to be associated with the rate of
CH~4~ flux than other broader or more shallowly conserved functions
[@schimel1998; @martiny2013]. This suggests that microbiome variation might
drive variation in CH~4~ oxidation in soils.

For the current project, we applied artificial ecosystem selection on soil
microbiomes by selecting on the ecosystem-scale CH~4~ oxidation rate in order
to estimate the amount of microbiome variation underlying variation in soil CH~4~ oxidation rate
[@swenson2000]. To identify which microorganisms could be markers of CH~4~
oxidation rate, we compared the composition of the artificially selected
microbiomes to a control set of microbiomes without selection. Artificial
ecosystem selection has a similar effect to enrichment culturing by amplifying
the population of interest [@swenson2000]. This will reduce the diversity of
the soil microbiome and allow for greater power in detecting significant
markers of CH~4~ oxidation rate. We then evaluated whether these markers meet
our underlying assumptions about which taxa limit the rate of ecosystem
function.

# Materials and Methods

## Experimental design

We performed an artificial ecosystem selection experiment [sensu @swenson2000]
by passaging replicate soil microbiomes. The trait we selected on was
CH~4~ oxidation rate. This experiment had two selection lines with twelve jars
each for a total of twenty-four jars per passage. One line was a positive
selection line where the two or three jars with the highest CH~4~ oxidation rate
were chosen to inoculate the next set of positive jars. The other line was a
neutral selection line where an equal number of jars as the positive line were
chosen at random to inoculate the next set of neutral jars. The number of jars
chosen was based on the distribution of fluxes among the positive jars, i.e., we
chose the top three jars unless only two jars had considerably greater CH~4~ oxidation
rates based on visual inspection of histograms. The experiment was carried out
over five passages until a significant divergence in functional rates was 
observed between the two selection treatments.

The initial soil microbiome was sampled from the top 10 cm of an upland mineral
soil under a deciduous forest ecosystem near the University of Oregon campus in
Eugene, OR, USA. Incubations were performed in 500 mL mason jars with rubber
septa installed in the lids. Each jar was sterilized with 70\% ethanol to which
was added 45 g of autoclaved potting mix, 5 g of living soil, and 3.5 mL of
sterile deionized water to bring them to 60\% of field capacity. The potting mix was
Lane Potting Mix from Lane Forest Products (Eugene, OR, USA). The jars were
then homogenized, capped, and injected with 4.3 mL of 99\% CH~4~ to bring
the headspace concentration to approximately 1000 ppm CH~4~. To create the
two treatment groups, twenty-four jars were created in an identical manner and
then randomly assigned to either the positive or neutral selection line. Jars
were flushed and respiked twice per week to maintain aerobic conditions and
elevated CH~4~ concentrations and were incubated at ambient temperature for
approximately three weeks. Methane oxidation rates were determined at the end
of the incubation period. For the positive treatment, the three jars with the
greatest CH~4~ oxidation rates were chosen to inoculate the next generation.
For the neutral treatment, three jars were randomly selected to inoculate the
next generation. For each treatment, these three jars were homogenized and 5 g 
of this homogenized soil was used to inoculate the next set of jars and which
represents a 10% subsampling. The
next set of jars were created in an identical manner to the first generation
with fresh autoclaved potting mix and the same moisture and CH~4~ content.

## Methane oxidation rate

Methane oxidation rates were determined after flushing and spiking jars to
approximately 1000 ppm CH~4~. Headspace samples of 1 mL were collected from each
jar immediately after spiking and then at time points 3, 6, 24, and 48 hours for
a 5-point curve. Samples were immediately injected into a SRI model 8610C gas
chromatograph equipped with a flame ionization detector (SRI Instruments, Torrance, CA, USA) to determine the headspace CH~4~
concentration. Fluxes were calculated from a first-order exponential decay
function as decay constant $k$ with units day^-1^. Oxidation rates are presented
as the additive inverse of $k$ (i.e., $-k$) so that a more positive value
represents a greater oxidation rate. The jars selected for the positive
treatment in passage 2 had the lowest CH~4~ oxidation rate of the twelve jars
due to a calculation error in the CH~4~ oxidation rate. All other passages
correctly used the top three jars.

A subsample of soil from the starting inoculum and from every jar in passages 2
and 5 was collected and stored at $-80^\circ$C. Soil DNA was extracted from 0.25
g soil. Negative controls were extracted from autoclaved potting mix and
DNase-free water. Extractions were performed using the DNeasy PowerSoil kit
(QIAGEN, Düsseldorf, Germany) and quantified using Qubit dsDNA HS Assay Kit (Thermo Fisher
Scientific, Inc., Waltham, MA, USA). To estimate the diversity and relative abundance of
the bacterial and archaeal taxa in our soil ecosystems, we sequenced the V4
region of the 16S rRNA gene using the using the 515F - 806R primer combination
[@caporaso2011]. PCR mixtures were: 10 $\mu$l NEBNext Q5 Hot Start HiFi PCR
master mix, 9.2 $\mu$l primer mixture (1.09 $\mu$M concentration), and 0.8
$\mu$l of DNA template. Reaction conditions were: $98^\circ$C for 30 s
(initialization); 35 cycles of $98^\circ$C for 10 s (denaturation), $61^\circ$C
for 20 s (annealing), and $72^\circ$C for 20 s (extension); and $72^\circ$C for
2 m (final extension). Reactions were performed in triplicate and then combined.
Amplicons were purified twice using 0.8x ratio Mag-Bind
RxnPure Plus isolation beads (Omega Bio-Tek, Norcross, GA, USA). Sequencing libraries were
prepared using a dual-indexing approach [@fadrosh2014; @kozich2013]. Amplicon
concentrations were quantified using Qubit and multiplexed at equimolar concentration. Sequencing
was performed at the University of Oregon Genomics Core Facility on the Illumina
NovaSeq 6000 with paired-end 150 bp reads (Illumina, Inc., San Diego, CA, USA).

Bioinformatics processing was performed in 'R' [@rcoreteam2018].
Demultiplexed sequencing reads were denoised using 'DADA2' to generate a table
of amplicon sequence variants (ASVs)
[@callahan2016]. Taxonomic assignment for the 16S reads was performed
using the Ribosomal Database Project naive Bayesian classifier [@wang2007]. The presence of
contaminants was evaluated using both the prevalence and frequency methods from
'DECONTAM' by comparing samples to negative controls [@davis2018]. We did not identify
any obvious contamination using these methods.

Statistical analyses were performed in 'R' [@rcoreteam2018]. To test
whether there was a significant change in CH~4~ oxidation rate ($k$) as a
response to selection, we tested a difference in slopes between the positive
and neutral selection lines. The CH~4~ oxidation rates were strongly 
right-skewed with most values close to zero and few large, positive values. This
resulted in residuals that did not meet the assumptions of constant variance
and normal distribution. Therefore, CH~4~ oxidation rates were log~10~
transformed to better meet the assumptions of a linear model and to make
figures easier to interpret. First, we tested the effect of treatment by
fitting two nested models with and without treatment using 'lm'. We compared these
models using the likelihood ratio test with the 'anova' function. We
then fit the full model to estimate the slope of the positive line, which
represents the change in CH~4~ oxidation rate per passage as a response to
selection.

To estimate the proportion of variance in ecosystem function due to variation in
the microbiome, we estimated narrow-sense heritability ($h^2$) as the
regression of mid-offspring on mid-parent [@falconer1996]. The mid-parent was the
mean for all three selected jars and the mid-offspring was the mean for all jars
produced by those parents. First, we tested if there was an effect of treatment
on the heritability estimate. We compared these models using the likelihood ratio
test with the 'anova' function. We then fit the full model to estimate the
heritability of the Positive and Neutral treatments.

Richness estimates and tests were performed using 'breakaway' [@willis2015]. We
tested a difference in richness between Passages 2 and 5 with both treatments
combined. Then we subset the samples from Passage 5 and compared richness between
the Positive and Neutral treatment. Next, we compared beta-diversity as both
Aitchison dissimilarity, which is the Euclidean distance after center-log ratio
transformation, and the binary Jaccard distance for presence or absence [@aitchison1982]. 
We tested a difference in centroid and dispersion of beta
diversity between the Positive and Neutral treatment and Passage 2 and 5 with
permutational analysis of variance (PERMANOVA) using the 'adonis2' function from
'vegan' [@mcardle2001; @oksanen2019].

To identify taxa that responded to selection on CH~4~ oxidation rate, we tested
differential abundance using a beta-binomial model [@martin2020]. We first
grouped ASVs at the family level keeping ASVs that lacked a family-level taxonomic
assignment. We
then subset the samples in Passage 5 and removed all families with a prevalence
of less than 10\%. We compared families in passage 5 between the Positive and
Neutral treatment to identify taxa that were enriched or depleted as a response
to selection. we used the likelihood ratio test to estimate p-values with a
significance threshold of false-discovery rate < 0.05 [@martin2020]. Estimates
are presented as the expected relative abundance in the positive treatment
relative to the neutral treatment.

We observed a response to selection on whole-ecosystem soil CH~4~ oxidation
rate (`r figures("response", display = "cite")`). In addition, the
response to selection varied with treatment 
`r print_lrt(compare_response_mods, 2, 'LRT of nested models with and without treatment')`. At the start of the experiment, the Positive
treatment had a mean CH~4~ oxidation rate that was
`r percent_less(initial = tidy(response_model)[[1, 2]], final = tidy(response_model)[[3, 2]] + tidy(response_model)[[1, 2]])`\% lower than the
Neutral treatment `r print_lm(response_model, 3, 'difference of y-intercepts')`. 
There was no change in CH~4~ oxidation rate of the Neutral
treatment over the five passages `r print_lm(response_model, 2)`. By contrast,
the Positive treatment had a `r round(per_slope)`\% increase in CH~4~
oxidation rate per passage `r print_lm(response_model, 4)`.

```{r response, fig.width=4, fig.height=3}
ratio_log10 <- calc_plot_ratios(fluxes$passage, fluxes$log_ch4)

p <- plot_flux(fluxes, passage, log_ch4, ratio_log10, log10 = TRUE)

ggsave('fig1.jpg', p)

@@

We estimated heritability as the regression of mid-offspring on mid-parent (`r figures("herit", display = "cite")`).
Offspring CH~4~ oxidation rates were correlated with parental CH~4~ oxidation
rates in both the Positive treatment `r print_lm(herit_model, 4)` and the
Neutral treatment `r print_lm(herit_model, 2)`. Notably, the sign of the effect
was reversed with a positive heritability for the Positive treatment and a negative heritability for the Neutral treatment.

```{r herit, fig.width=5, fig.height=3}

p <- plot_herit(heritability)

ggsave('fig2.jpg', p)

@@

Sequencing of 16S rRNA amplicons revealed that there were `r ntaxa(ps)` unique
ASVs across all `r nsamples(ps)` jars sampled.
Richness of ASVs decreased among all of the jars between Passage 2 and 5
(Richness difference between Passage 5 and Passage 2 = `r myround(bt_passage$table[2, 1], 1)`, SE =
`r myround(bt_passage$table[2, 2], 2)`, p < 0.001). However, there was no
difference in richness between the Positive and the Neutral treatment in
Passage 5 (Difference = `r myround(bt_treat5$table[2, 1], 1)`, SE =
`r myround(bt_treat5$table[2, 2], 2)`, p =
`r myround(bt_treat5$table[2, 3], 3)`).  In addition, there was no
correlation between richness and CH~4~ oxidation rate across all passages
and treatments.  (slope = `r myround(bt_ch4$table[2, 1], 1)`, SE =
`r myround(bt_ch4$table[2, 2], 2)`, p = `r myround(bt_ch4$table[2, 3], 3)`).

Taxonomic dissimilarity of the soil microbiome varied strongly by passage and
weakly by treatment (`r figures("beta", display = "cite")`). Permutational
analysis of variance demonstrated an effect of passage, treatment,
and their interaction on Aitchison dissimilarity (`r tables("adonis", display = "cite")`). 
Passage explained `r myround(100*fit_adonis[1, 4])`\% of the
variation in Aitchison dissimilarity and `r myround(100*fit_adonis[6, 4])`\% in
Jaccard dissimilarity.  Treatment explains `r myround(100*fit_adonis[2, 4])`\%
for Aitchison and `r myround(100*fit_adonis[7, 4])`\% for Jaccard.  Variation
in microbiome composition decreased as a result of passaging in jars (F = 80.5,
p < 0.001), but did not differ between selection treatments (F = 0.40, p =
0.54).

```{r beta, results=FALSE, fig.height=2.5}

p <- plot_beta(ps)

ggsave('fig3.jpg', p)

```{r adonis}

write.table(fit_adonis, 'table_1.tsv')
# kable(col.names = c('Term', 'df', 'SS', 'R2', 'F', 'p'), 
#       digits = c(2,2,2,2,2,3),
## 	format = "Table",
## 	booktabs = T,
# 	caption = "PERMANOVA on Bray-Curtis and Jaccard dissimilarities.",
#   	label = "beta") %>%
#   pack_rows("Bray-Curtis", 1, 5) %>%
#   pack_rows("Jaccard", 6, 10)

To investigate which taxa responded to selection on soil CH~4~ oxidation rate, we
first looked at which taxa were unique to the Positive and Neutral treatments
aggregated at the family level. Several taxa were present in the Positive
treatment and absent in the Neutral treatment. Most of these families had low
prevalence (present in 3 or fewer samples) and low abundance (median < 5 reads).
There were two families unique to the Positive treatment with relatively high
prevalence. This included an ASV that was a member of the Bacteroidia Class with
no lower taxonomic designation. This ASV had a prevalence of 10/12 and a median
abundance of ~2 reads. The other prevalent family was a member of the
Silvanigrellaceae, a newly described family placed in its own order. This family
was present in all 12 samples and had a median abundance of 8 reads.
Silvanigrella is the only cultivated member of
Silvanigrellaceae and was isolated from a temperate fresh water lake
[@hahn2017]. Of the families unique to the Neutral treatment, only one had a
prevalence greater than 2/12. This family, Armatimonadaceae, was present in half
of the Neutral samples (prevalence = 6/12). The type strain for Armatimonadaceae was isolated from
the rhizosphere of *Phragmites australis* [@tamaki2011].
